### 2021-06-14 abstractESSLLI21/abstractESSLLI1.Rmd
revised abstract

To improve on the situation, we  build  Bayesian models to estimate the biases involved using the raw datapoints, actually using control groups,  distinguishing the connection types,  and taking the uncertainty involved seriously.
For the general impact of being associated, we build  models using STAN according to the following specification (linear categorical model and the priors):

@@

\begin{align}
cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
\mu_i & = m_{pw} + co_{con}\\
m_{pw} & \sim dnorm(1,.5) \\
co_{con} & \sim dnorm(0,1) \\
\sigma &\sim  dcauchy(0,1)
\end{align}
\normalsize

The resulting coefficients for the religion dataset based on Reddit embeddings are in Figure 5. While there is some difference in the means, the 89\% highest posterior density invervals are quite wide and include 0s for all the coefficients.  Then, motivated by large differences between the states of different protected words, we build bayesian models with separate mean estimates, with the same regularizing priors as before. The results for the religion dataset based on Reddit embeddings are in Figure 6.

We build analogous models for other topic groups (race, gender), and different embeddings (Google, and hard-debiased Reddit). The general conclusion is that once we proceed this way, the situation is much less obvious. The word list sizes are small, sample sizes are small, and so posterior density intervals are wide. Some stronger bias can be discerned in the gender case. Moreover, sometimes the differences between associated, different and human, are not very impressive. The cost of debiasing turns out sometimes to be that   neutral human predicates get closer to protected classes (in religion),  change in gender after debiasing is really minor with   zero still out out of HPDI range, and debiasing improvement for race is a small improvement   achieved at the price of moving neutral terms closer to protected words (there is no space to visualise these results in this abstract).

@@

ggplot()+geom_histogram(aes(x=biasesD2, y = ..density..), alpha = 0.6, bins=50)+
  theme_tufte()+labs(title="Figure 2.", subtitle = "10k biases for different means and sd =.05")+ xlab("bias")
```
   \end{minipage}

>>>>

To improve on the situation, we  build  Bayesian models to estimate the biases involved using the raw datapoints, actually using control groups,  distinguishing the connection types,  and taking the uncertainty involved seriously.
For the general impact of being associated, we build  models using hamiltonian Monte Carlo (STAN) according to the following specification (linear categorical model and the priors):

@@

\begin{align}
cosineDistance_i  & \sim dnorm(\mu_i, \sigma) &
\mu_i & = m_{pw} + co_{con} \tag{Model}\\
m_{pw} & \sim dnorm(1,.5)  & 
co_{con} & \sim dnorm(0,1)  \,\,\,\,\,
 & \sigma  \sim  dcauchy(0,1) \tag{Priors}
\end{align} 
\normalsize

\noindent That is, given a protected word we assume its cosine distance to  attributes from a given group is normally distributed around a mean, which is determined by the mean assigned to this protected word and by the connection type of that group of attributes (stereotypically associated/ associated with a different predicate/neutral human attribute/random word), and we use weak regularizing priors for the parameters. The task of the model is to estimate the mean for a given protected word and the coefficients of change in that mean corresponding to different connection types.

The resulting coefficients for the religion dataset based on Reddit embeddings are in Figure 5. While there is some difference in the means, the 89\% highest posterior density invervals are quite wide and include 0s for all the coefficients.  Then, motivated by large differences between the states of different protected words, we build bayesian models with separate coefficients for protected classes, with analogous regularizing priors. The results for the religion dataset based on Reddit embeddings are in Figure 6.

We build analogous models for other topic groups (race, gender), and different embeddings (we use  three embeddings: GoogleNews-vectors-negative300, and Reddit Corpus, and the hard-debiased Reddit embedding used by @Manzini2019blackToCriminal). The general conclusion is that from this bayesian perpective the situation is much less obvious. The word list sizes are small, sample sizes are small, and so posterior density intervals are wide. Some stronger bias can be discerned in the gender case. Moreover, sometimes the differences between associated, different and human, are not very impressive. The cost of debiasing turns out sometimes to be that   neutral human predicates get closer to protected classes (in religion),  change in gender after debiasing is really minor with   zero still out out of HPDI range, and debiasing improvement for race is a small improvement   achieved at the price of moving neutral terms closer to protected words (there is no space to include these results in this abstract).

@@

ggplot()+geom_histogram(aes(x=biasesD2, y = ..density..), alpha = 0.6, bins=50)+
  theme_tufte()+labs(title="Figure 2.", subtitle = "10k biases for different means  and sd =.05 (case 3 in the table)")+ xlab("bias")
```
   \end{minipage}

### 2021-06-14 abstractESSLLI21/abstractESSLLI1.tex
revised abstract

\begin{align}
cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
\mu_i & = m_{pw} + co_{con}\\
m_{pw} & \sim dnorm(1,.5) \\
co_{con} & \sim dnorm(0,1) \\
\sigma &\sim  dcauchy(0,1)
\end{align}

@@

The resulting coefficients for the religion dataset based on Reddit
embeddings are in Figure 5. While there is some difference in the means,
the 89\% highest posterior density invervals are quite wide and include
0s for all the coefficients. Then, motivated by large differences
between the states of different protected words, we build bayesian
models with separate mean estimates, with the same regularizing priors
as before. The results for the religion dataset based on Reddit
embeddings are in Figure 6.

We build analogous models for other topic groups (race, gender), and
different embeddings (Google, and hard-debiased Reddit). The general
conclusion is that once we proceed this way, the situation is much less
obvious. The word list sizes are small, sample sizes are small, and so
posterior density intervals are wide. Some stronger bias can be
discerned in the gender case. Moreover, sometimes the differences
between associated, different and human, are not very impressive. The
cost of debiasing turns out sometimes to be that neutral human
predicates get closer to protected classes (in religion), change in
gender after debiasing is really minor with zero still out out of HPDI
range, and debiasing improvement for race is a small improvement
achieved at the price of moving neutral terms closer to protected words
(there is no space to visualise these results in this abstract).

>>>>

\begin{align}
cosineDistance_i  & \sim dnorm(\mu_i, \sigma) &
\mu_i & = m_{pw} + co_{con} \tag{Model}\\
m_{pw} & \sim dnorm(1,.5)  & 
co_{con} & \sim dnorm(0,1)  \,\,\,\,\,
 & \sigma  \sim  dcauchy(0,1) \tag{Priors}
\end{align}

@@

\noindent That is, given a protected word we assume its cosine distance
to attributes from a given group is normally distributed around a mean,
which is determined by the mean assigned to this protected word and by
the connection type of that group of attributes (stereotypically
associated/ associated with a different predicate/neutral human
attribute/random word), and we use weak regularizing priors for the
parameters. The task of the model is to estimate the mean for a given
protected word and the coefficients of change in that mean corresponding
to different connection types.

The resulting coefficients for the religion dataset based on Reddit
embeddings are in Figure 5. While there is some difference in the means,
the 89\% highest posterior density invervals are quite wide and include
0s for all the coefficients. Then, motivated by large differences
between the states of different protected words, we build bayesian
models with separate coefficients for protected classes, with analogous
regularizing priors. The results for the religion dataset based on
Reddit embeddings are in Figure 6.

We build analogous models for other topic groups (race, gender), and
different embeddings (we use three embeddings:
GoogleNews-vectors-negative300, and Reddit Corpus, and the hard-debiased
Reddit embedding used by {[}6{]}). The general conclusion is that from
this bayesian perpective the situation is much less obvious. The word
list sizes are small, sample sizes are small, and so posterior density
intervals are wide. Some stronger bias can be discerned in the gender
case. Moreover, sometimes the differences between associated, different
and human, are not very impressive. The cost of debiasing turns out
sometimes to be that neutral human predicates get closer to protected
classes (in religion), change in gender after debiasing is really minor
with zero still out out of HPDI range, and debiasing improvement for
race is a small improvement achieved at the price of moving neutral
terms closer to protected words (there is no space to include these
results in this abstract).
